E. coli Testing: Safeguarding Public Health

### 2024-09-27 paper/paper.qmd
Update the document

The increase in E. coli concentration is a serious problem for humans. Elevated concentrations make beach water unsafe, hindering safe human use. Although some strains of E.coli are harmless, pathogenic microorganisms may still be present. In cases of increased concentration, healthy individuals among those infected can recover on their own; however, vulnerable populations such as children, the elderly, and those with compromised immune systems may suffer more severe illnesses, including general gastrointestinal discomfort, rashes, ear and eye infections, and in severe cases, bloody diarrhea, kidney failure, and death. The rise in E.coli concentration is caused by various factors, such as the runoff of animal feces carrying pathogens from land or sewage systems, aging sewage pipelines, and wastewater infrastructure issues. Considering the rapid growth of E. coli and the associated public health risks, real-time monitoring of beach water quality is essential.

According to the requirements of the Ministry of Health Beach Management Protocol (January 1, 1998), the Beach Water Sampling Program for the City of Toronto was implemented to reduce the incidence of waterborne diseases. Toronto's beaches have been certified under the Blue Flag program. From June to September each year, the relevant departments in Toronto collect water samples daily from all regulated beaches in the city to test for E.coli bacteria concentrations.

The number of E.coli in freshwater is determined by counting the number of yellow and yellow-brown colonies that grow after placing a 0.45-micron filter on m-TEC medium and incubating it at 35.0ºC for 22-24 hours @citebacteria. The water quality standard for E.coli in Ontario and federally is 200 E.coli per 100 milliliters of water, while Toronto beach water quality standard is 100 E.coli per 100 milliliters of water @citewaterquality. This article utilizes 22,000 data points collected by government departments from June to September from 2007 to 2024 to assess whether the E.coli is much more than the Toronto water quality standard. Meanwhile, it try to prove if the E.coli is influenced by the change of time and site location.

@@

## The time influence E.coli
From a time perspective, based on the information provided in the previous section's graphs and tables, there have been significant changes in Toronto beach water quality between 2007 and 2024. Particularly in 2008, E.coli levels peaked, with an average concentration approaching 300 E.coli/100ml. After that, the E.coli levels showed a downward trend and became more stable. This result may be related to several factors mentioned in the introduction, such as animal feces, human activities, and the aging of the storm water drainage system. Based on the evidence provided by the trend in Figure 3, there is a clear negative correlation between the average E.coli levels and time. However, since the average value alone cannot be used as a final indicator to judge the trend, we can refer to the percentage of exceedance bars in @fig-histogram-proportion. The exceedance rate of E.coli in 2008 exceeded 50%, indicating that water quality from June to September of that year was poor, leading to reduced activities and potential impacts on health and hygiene. In contrast, 2016 had the lowest E.coli levels, being closest to Toronto beach water quality standard of 100 E.coli/100ml, with an exceedance rate of only about 20%. This may be related to the government's efforts to improve the beach environment and favorable weather conditions from June to September. Data from 2020 was limited due to collection restrictions caused by the COVID-19 pandemic, but there was no significant increase compared to other years, indicating that the overall water quality remained stable. Over time, both graphs show that after two peaks, E.coli levels have generally declined and stabilized.

## Whether the Site location influence E.coli
In addition to time factors, evaluating whether the test locations affect E.coli levels can further reveal water quality differences across various measurement sites. Based on @fig-scatter-mean-site, we can clearly see that the average E.coli levels vary significantly depending on the test location. E.coli levels at Sunnyside Beach are generally lower, indicating relatively better water quality in this area, especially at the 20W test site, where the exceedance rate is only 35.08866%, according to @tbl-proportion-site. On the other hand, at the Marie Curtis Park East Beach, E.coli levels are relatively higher, particularly at the 29W and 33W test sites, where exceedance rates are close to 40%. However, although differences exist between test sites, the average levels at all sites still exceed Toronto water quality standards, though they are below Canadian water quality standards. Since average concentration alone cannot fully determine whether the water quality of an area is non-compliant, the calculation of exceedance rates in Table 2 helps reveal that only a small proportion of water quality is non-compliant. The high exceedance values contribute to the elevated averages. By calculating exceedance rates by test location, the analysis becomes more insightful and evidence-based, providing more convincing results.

## Conclusion
Overall, through the analysis of E.coli data from 2007 to 2024 in Toronto beaches, we have uncovered trends in concentration changes over time and across different locations. While there has been an improvement in overall water quality, some years and measurement sites still exhibit high E.coli concentration exceedances. Additionally, the influence of time on E.coli levels cannot be fully confirmed due to the variability of factors other than data and the uncertain governmental interventions.

>>>>

The increase in E.coli concentration poses a serious risk to human health. Elevated levels render beach water unsafe, impeding recreational use. While some strains of E.coli are harmless, pathogenic microorganisms may still be present. In cases of heightened concentration, healthy individuals can often recover on their own; however, vulnerable populations—such as children, the elderly, and those with compromised immune systems—are at greater risk of severe illnesses, including gastrointestinal discomfort, rashes, ear and eye infections, and, in extreme cases, bloody diarrhea, kidney failure, and even death. The rise in E.coli concentration is attributed to various factors, including runoff from animal feces carrying pathogens, aging sewage pipelines, and issues with wastewater infrastructure. Given the rapid growth of E.coli and the associated public health risks, real-time monitoring of beach water quality is essential.

In accordance with the Ministry of Health’s Beach Management Protocol, the Beach Water Sampling Program for the City of Toronto was implemented to reduce the incidence of waterborne diseases. Toronto’s beaches are certified under the Blue Flag program. From June to September each year, relevant departments collect water samples daily from all regulated beaches in the city to test for E.coli concentrations.

E.coli levels in freshwater are determined by counting the yellow and yellow-brown colonies that grow after applying a 0.45-micron filter on m-TEC medium and incubating it at 35.0ºC for 22-24 hours @citebacteria. The water quality standard for E. coli in Ontario and federally is 200 E.coli per 100 milliliters of water, while Toronto’s beach water quality standard is 100 E.coli per 100 milliliters of water @citewaterquality. This article utilizes 22,000 data points collected by government departments from June to September between 2007 and 2024 to assess whether E.coli levels exceed Toronto’s water quality standard. Additionally, it aims to determine if E.coli concentrations are influenced by changes over time and variations in site location.

@@

## The time influence E.coli
From a temporal perspective, the data presented in the previous section's graphs and tables indicate significant changes in the water quality of Toronto's beaches from 2007 to 2024. Notably, in 2008, E.coli levels reached a peak, with average concentrations approaching 300 E.coli per 100 ml. Subsequently, E.coli levels exhibited a downward trend and became more stable. This improvement may be attributed to several factors mentioned in the introduction, including animal waste, human activities, and the aging stormwater drainage system.

The trend illustrated in Figure 3 shows a clear negative correlation between average E.coli levels and time. However, since the average value alone cannot serve as a definitive indicator of this trend, we can also consider the exceedance rates displayed in @fig-histogram-proportion. In 2008, the exceedance rate of E. coli surpassed 50%, indicating poor water quality from June to September of that year, which likely curtailed recreational activities and posed potential health risks. In contrast, 2016 marked the lowest levels of E.coli, aligning closely with the Toronto beach water quality standard of 100 E.coli per 100 ml, with an exceedance rate of only about 20%. This improvement may be linked to government initiatives aimed at enhancing the beach environment, as well as favorable weather conditions during the summer months.

Data collection in 2020 was limited due to restrictions imposed by the COVID-19 pandemic; however, there was no significant increase in E.coli levels compared to other years, suggesting that overall water quality remained stable. Over time, both graphs demonstrate that after experiencing two peaks, E.coli levels have generally declined and stabilized.

## Whether the Site location influence E.coli
In addition to temporal factors, evaluating whether testing locations impact E.coli levels can further highlight differences in water quality across various measurement sites. As illustrated in @fig-scatter-mean-site, average E.coli levels vary significantly based on the test location. Sunnyside Beach generally exhibits lower E.coli levels, indicating relatively better water quality, particularly at the 20W test site, where the exceedance rate is only 35.09%, according to @tbl-proportion-site. In contrast, Marie Curtis Park East Beach shows relatively higher E. coli levels, especially at the 29W and 33W test sites, where exceedance rates approach 40%.

Despite these differences among test sites, average E.coli levels at all locations still exceed Toronto's water quality standards, although they remain below Canadian water quality standards. Since average concentration alone cannot fully determine whether an area’s water quality is compliant, the exceedance rates calculated in @tbl-proportion-site provide additional clarity, revealing that only a small proportion of water quality measurements fall outside compliance. The elevated exceedance values contribute to the higher average levels. By calculating exceedance rates by test location, the analysis becomes more insightful and evidence-based, yielding more compelling results.

## Conclusion
Overall, the analysis of E.coli data from Toronto beaches between 2007 and 2024 has revealed trends in concentration changes over time and across various locations. While there has been a notable improvement in overall water quality, certain years and measurement sites still exhibit significant exceedances in E.coli concentrations. Moreover, the influence of time on E.coli levels cannot be fully confirmed due to the variability of factors beyond the data and the uncertainties surrounding governmental interventions.
